---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
print("Hello puppy!")
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example:

```{r}
setwd("~/code/data_journalism_2023_spring")
```
### What Is The Tidyverse?
###library(tidyverse)
```

The library function loads an R library such as the tidyverse, providing access to the verbs and functions that library contains. You have to load a library before you can use its functions.

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

### How to read in a CSV file from the Internet

```{r}
wes_moore_contributions <- read_csv("https://raw.githubusercontent.com/stephenneukam/CNS_Annapolis/main/Campaign_finance/Moore_ContributionsList.csv")
```

### How to count rows

You use the `n()` function to count rows, like so:

```{r}
dataframe %>% 
  group_by(some_column) %>% 
  summarize(count = n())
```
### Group and count

The keyboard shortcut for this is cmd-shift-m on the Mac and ctrl-shift-m on Windows. Don't like that character? R now has a new one that does the same thing: `|>`.

Group by is a great way to see if you have bad data because it shows you each group (e.g. monday, tuesday, wednesday, blue)

Summarise does all the math stuff - you group by and use summarize to count, add, subtract, multple, etc., "n" does the math - counts the number of rows
```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
### Glossary of Examples

###**Task** Run the codeblock below to load the data, which is contained in an RDS file.

umd_courses <- read_rds("lab_01.rds")


###Filter and Select
journalism_courses <- umd_courses |> filter(columname == "criteria")
The following filters for terms after 2023 (no quotes with numbers)
courses_2023 <- umd_courses |> filter(term > 202300)

head(courses_2023) 

Filter that department or whatever subset of data (data frame), then run other things on it (e.g. summarise etc)

Select shows you specific columns, gets rid of other columns (basically hides them)

can do select(courses, instructor) or can do select(-syllabus_count) to hide everything except syllabus count, can also select all columns that start with these three characters

To COMBINE filters: 
theatre_seats_15 <- umd_courses |> filter(department == "Theatre") |> filter(seats >= 15)

nrow(theatre_seats_15) OR can combine:
and_theatre_seats_15 <- umd_courses |> filter(department == "Theatre", seats >= 15) 
Comma does both filters together, | means Either theater OR courses with 15 seats
doing != does "department does not equal theater"
nrow(and_theatre_seats_15)

### Task 7: Filter on partial match
umd_courses |>
  filter(str_detect(title, "Shakespeare")) brings back any title that has "shakespeare" in it
  The latest version of the tidyverse also has a function called `str_like` which uses the wildcard search operator % to do partial matching.
  
  ### MUTATE
convert date column with characters into real dates
maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))

head(maryland_expenses) 
need the <- to make the change permanent (make this change then save it)