---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
print("Hello puppy!")
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example:

```{r}
setwd("~/code/data_journalism_2023_spring")
```
### What Is The Tidyverse?
###library(tidyverse)
```

The library function loads an R library such as the tidyverse, providing access to the verbs and functions that library contains. You have to load a library before you can use its functions.

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

### How to read in a CSV file from the Internet

```{r}
wes_moore_contributions <- read_csv("https://raw.githubusercontent.com/stephenneukam/CNS_Annapolis/main/Campaign_finance/Moore_ContributionsList.csv")
```

### How to count rows

You use the `n()` function to count rows, like so:

```{r}
dataframe %>% 
  group_by(some_column) %>% 
  summarize(count = n())
```
### Group and count

The keyboard shortcut for this is cmd-shift-m on the Mac and ctrl-shift-m on Windows. Don't like that character? R now has a new one that does the same thing: `|>`.

Group by is a great way to see if you have bad data because it shows you each group (e.g. monday, tuesday, wednesday, blue)

Summarise does all the math stuff - you group by and use summarize to count, add, subtract, multple, etc., "n" does the math - counts the number of rows, 
```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
### Glossary of Examples
