---
title: "lab_10"
author: "Derek Willis"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Load libraries
library(tidyverse)
library(sf)
library(janitor)
library(tidycensus)
library(tigris)
```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Maryland voters made history last November by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?


**A1.**

Looks like most counties had a very similar percentage that voted for both Moore and Brown. The biggest differences were in Montgomery County, which favored Moore by more than 1 percent, Prince George's and Cecil counties, which favored Brown by less than half a percent, and Baltimore City which favored Brown by more than 1 percent.

```{r}
md_statewide_county <- read_csv("data/md_statewide_county.csv") |> clean_names()
md_statewide_county$geoid <- as.character(md_statewide_county$geoid)
counties <- counties() |> clean_names() |> rename(county = name)
  

md_statewide_county |> 
  summarise(sum(moore+cox+lashar+wallace+harding+sefcik+gov_writein))

#total_votes_gov = 1754475
#total_votes_ag = 1503462				

md_statewide_county |>
  summarise(sum(peroutka+brown+ag_writein))

md_statewide_county <- md_statewide_county |>
  mutate(pct_moore = moore/1754475*100, pct_brown = brown/1503462*100, difference = pct_moore-pct_brown)

md_elections_and_geometry <- counties |> inner_join(md_statewide_county, join_by(geoid))

```
```{r}
ggplot() +
  geom_sf(data=md_elections_and_geometry, aes(fill=difference)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma")
```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland from last March. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.**
Looks like what I believe is Talbot County experienced a significantly high number of power outages in March. I would like to create a different scale so I could get a better sense of what is happening between the 54 and 403 outages in the pink counties. My next steps would be to check to see what the average was, what the highest months for power outages were, and check the demographics. Then I'd look into what causes power outages and call the power company probably.

```{r}
county_outages_march23 <- read_csv("data/county_outages_march2023.csv")
md_counties <- counties |>
  filter(statefp == "24")

county_outages_march23 <- county_outages_march23 |>
  mutate(outages_per_10k = outage_customers/total_customers*10000)

outages_county_map <- md_counties |> inner_join(county_outages_march23, join_by(county))

ggplot() +
  geom_sf(data=outages_county_map, aes(fill=outages_per_10k)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log")
```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change. What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.**
Caroline and Dorchester counties stand out the most to me - they had the highest and lowest percent change in EV vehicles over the last 3 years. I'd be interested to see if counties or the state offer some kind of incentive for people to get EVs. I would also be interested to see how these numbers correlate with median income. 

Our second map looks much more like what I expected to see - areas closer to DC have more EVs per capita, and areas on the outer edges of MD have fewer. Montgomery and Howard counties have the most EVs per capita. Somerset has the fewest. I would still be interested in looking at this with median income data. I'm surprised Baltimore doesn't have more but maybe that has something to do with wealth?

Anyway it's different from the first map I think because it has more variation - it highlights bigger differences and shows us that, while Caroline County had the highest percent increase in EVs, they have 50 or less per capita, so they must have had few EVs to start with. Dorchester didn't see a high percentage increase and also doesn't have a ton of EVs per capita. Howard and Montgomery saw moderate gains compared to other areas, but have a high rate of EVs per capita overall.

```{r}
ev_data_county <- read_csv("data/ev_2020_2023.csv") |>
  mutate(pct_change = ((march_2023-july_2020)/july_2020)*100)

ev_county_map <- md_counties |> inner_join(ev_data_county, join_by(county))

md_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              year = 2021,
              state = "MD")

md_county_population <- md_county_population |>
  rename(county = NAME) |>
  mutate(county = str_to_title(str_remove_all(county,"County, Maryland"))) |>
  mutate(county = str_trim(county,side="both"))

ev_county_map_population <- ev_county_map |> inner_join(md_county_population, join_by(county))

ev_county_map_population <- ev_county_map_population |> 
  rename(population = estimate) |>
  mutate(ev_per_capita = march_2023/population*10000)

```

```{r}
ggplot() +
  geom_sf(data=ev_county_map, aes(fill=pct_change)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log")
```

```{r}
ggplot() +
  geom_sf(data=ev_county_map_population, aes(fill=ev_per_capita)) +
  theme_minimal()+
  scale_fill_viridis_b(option="magma")

```

