---
title: "lab_04"
author: "Taylor Nichols"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/data_journalism/data_journalism_fall_2023/labs/lab_04/data")
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

### How to bring data out of R studio - write_csv (google example)

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# load your packages
library(tidyverse)
library(lubridate)
library(janitor)

```

For this lab, we want to investigate spending by Maryland state and local candidates during the previous election. 

For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? 

We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table

setwd("~/data_journalism/data_journalism_fall_2023/labs/lab_04/data")

md_expenses <- read_csv("maryland_expenses.csv")

md_expenses
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

   1. Create a dataframe of Media expenses where the name of the payee is not NA.
   2. Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
   
   3. Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
   
   4. When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
   
   5. Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.**
Note: I used chatgpt for this, asked it "how do I use the filter function in R to find any rows where the value is not NA?" and it told me to use !is.na function with filter.

Friends of Jain Ashwani spent the most on Facebook and Meta advertising, with a total of $62,357.38.

```{r}
setwd("~/data_journalism/data_journalism_fall_2023/labs/lab_04/data")

md_expenses <- md_expenses |> 
  clean_names()

md_media_expenses <- md_expenses |>
  filter(expense_category == "Media") |>
  filter(!is.na(payee_name))

write_csv(md_media_expenses, "md_media_expenses.csv")

payee_clean <- read_csv("final_clean_maryland_media_expenses.csv")

```
```{r}
facebook_ads <- payee_clean |> filter (payee_clean == "FACEBOOK/META") |>
  group_by(committee_name) |>
  summarise(total = sum(amount))|>
  arrange(desc(total))

facebook_ads
```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\%5Bdirection\%5D=desc&sort_data\%5Bmode\%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?


**A2.** 
Most of these groups were for people who were running for governor. I looked up ads by Jain Ashwani / Friends of Jain Ashwani and they all were removed for violating Facebook's guidelines on election ads. Looking at the ads, I don't see anything that looks alarming. They're videos of him talking about when he had cancer, his ideas on tackling climate change, or that he's a voter-funded candidate. Seems worth looking more deeply into.


**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**A3.** 
Wes Moore is missing! A lot of other people are as well - all the Republican candidates, and the runner-ups to Moore, Tom Perez and Shannon Sneed. It doesn't appear that Perez/Sneed or Cox/Schifanelli ran facebook ads? that seems really weird to me. I also don't see Wes Moore in here so I'm suspicious that I may have missed something in my data cleaning?

Looks like they spent a LOT of money on SKDK which is a media company that manages communications for campaigns. My guess is that company subsequently paid FB and all other vendors but it just got logged in the data as SKDK. Not very helpful for us :)

```{r}
payee_clean |> filter (str_detect(committee_name, "moore|Moore|MOORE")) |>
  group_by(payee_clean) |>
  summarise(total = sum(amount)) |>
  arrange(desc(total))

```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** 
Overall this list looks pretty different from the first one. It does look like Peter Franchot's campaign spent a significant chunk of their online media funds on Facebook ads. Jain Ashwani's campaign spent all their online media funds on Facebook/Meta ads.

```{r}
payee_clean |> filter(expense_purpose == "Online Advertising") |> 
  group_by(committee_name) |>
  summarise(total = sum(amount)) |>
  arrange(desc(total))
  
```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5**
Well, we have some pretty significant gaps in our data because of funds that went to media orgs like SKDK, so we have absolutely no clue how much money Moore's campaign spent on Facebook. We also have gap in our data because Meta owns Facebook and Instagram so we can't compare ad dollars spent on FB vs IG. I think it would be interesting to look at how much revenue Facebook makes off of campaign advertising (not necessarily from just this dataset but would be worth looking into if we got a bunch of data). Anyways I think the most interesting story is that Jain Ashwani's facebook ads were hidden or removed for violating the facebook conditions. Would be worth looking into WHY they were taken down and whether or not other candidates had a similar experience. ALSO of note / worth looking into is where did the republicans spend their ad dollars?? I would imagine facebook would be a hot spot for them. 
